---
title: "Camparison of 401K VS Social Security for Retirement"
author: "Partha Banerjee, Rohan Fray, Vincent Ying"
date: "November 19, 2014"
output: html_document
---


  
# Library and Directory Config
```{r echo=FALSE}
# Add all required packages here. 
required.packages <- c("plyr", 
                       "ggplot2",
                       "gdata")

# Check to see if packages in required.packages are installed
have.pkgs <- required.packages %in% installed.packages()

if(length(required.packages) > sum(have.pkgs)){
  install.packages(required.packages[!have.pkgs])
}

for(pkg in required.packages){
  library(pkg, character.only = TRUE)
}

# Make sure we're in the right working directory:  "Code"
if("Code" %in% list.files()){
  setwd("Code")
}

if(!("Code" %in% list.files("..")))
{
  stop
}
```

# Load Data

```{r data}
# SSA Dataset (Taken from supplement13.xlsx)
ssa_taxrate = read.csv("../Data/OASDI - Table 2.A3 - Annual max taxable earnings and contribution rates.csv")
ssa_taxamt = read.csv("../Data/OASDI - Table 2.A4 - Maximum annual contribution.csv")
ssa_benefit = read.csv("../Data/OASDI - Table 6.B3 - All Benefit.csv")


age <- c(62, 63, 64, 65, 66, 67, 68, 69, 70)

# SSA - Average montly benefit from 2012
#   supplement13, Table 6.B3 
#   and http://www.ssa.gov/OACT/ProgData/ar_drc.html
mth_amt <- c(1103.9, 1182.75,	1261.6,	1366.62, 1471.81,	1577, 1703.16, 1829.32, 1955.48)
ssa_benefit <- data.frame(age, mth_amt)

# 401k - Base mount accumulated from contribution
#   https://blog.personalcapital.com/financial-planning-2/average-401k-balance-age/
tot_low <- c(690500, 708000, 725500, 743000,	760500, 778000, 795500, 813000, 830500)
# 401k - Balance with 7% compound interest
#   http://www.investor.gov/tools/calculators/compound-interest-calculator
tot_high <- c(3613402, 3883840, 4173209, 4482834,  4814132, 5168621, 5547925, 5953779, 6388044)
k_Balance <- data.frame(age, tot_low, tot_high)
```


# Comparison of SSA and 401k return.

Assumptions for comparison:
- Start work at age 22.
- Life Expectency for birth in 1991 is 75.8 years.
- Use average SSA benefit from 2012
- Start contributing to 401k from age 22 with contributions the same amount as SSA contribution.
- Use 7% compound growth for 401k contributions.


```{r}


```


# Modeling:
Problem:
- When to start retirement?
- When to start withdrawing from SSA?


Assumptions for modeling problem:
- Start work at age 22.
- Average life expectency till 85.45 for those reaching 65 in 2014.
- Using average SSA benefit from 2012
- Start contributing to 401k from age 22 with an inital contribution of $8,000 and $17,500 every year after.
- Use base amount of 401k contributions.


Models:
A. Optimization of Discrete Models with Simplex Method
```{r}
# Calculating SSA Total Benefit with Average Death at 85.45
ssa_benefit$tot_amt <- (85.45-ssa_benefit$age) * 12 * ssa_benefit$mth_amt

# Calculate 401k Monthly Benefit with Base Balance
k_Balance$mth_low <- k_Balance$tot_low / (85.45-k_Balance$age) / 12
# Calculate 401k Monthly Benefit with 7% Increase
k_Balance$mth_high <- k_Balance$tot_high / (85.45-k_Balance$age) / 12

# Combined benefit when withdrawing 401k at 62 and delaying SSA
tot_amt <- ssa_benefit$tot_amt + k_Balance[k_Balance$age==62,]$tot_high
mth_amt <- tot_amt / (85.45 * 12)
combin <- data.frame(age, mth_amt, tot_amt)

# Combined benefits from SSA and 401k
plot(combin$age, combin$mth_amt)
plot(combin$age, combin$tot_amt)


# Combined benefit when contributing to 401k until start of SSA
tot_amt <- ssa_benefit$tot_amt + k_Balance$tot_high
mth_amt <- tot_amt / (85.45 * 12)
combin_401contrib <- data.frame(age, mth_amt, tot_amt)
```

B. Modeling with Decision Theory
```{r}


```