---
title: 'ISDA 609 - Mathematical Modeling Techniques for Data Analytics'
author: "Partha Banerjee, Rohan Fray, Vincent Ying"
date: "November 19, 2014"
output:
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 4
  html_document:
    theme: united
    toc: yes
geometry: margin=0.6in
fontsize: 11pt
---

\pagebreak

```{r clear, include=FALSE, echo=FALSE}
# Added this "clear all" section to make sure we can run use "Chunks/Run All"" without conflict.
# Still does not catch package info, but that will be seen when knit.

# Start with a Clean memory
rm(list=ls())

# Supress warnings particularly during loading of libraries
options(warn=-1)
```

```{r config, echo=FALSE, results='hide'}
# Library and Directory Config

# Add all required packages here. 
required.packages <- c("plyr", 
                       "ggplot2")

# Check to see if packages in required.packages are installed
have.pkgs <- required.packages %in% installed.packages()

if(length(required.packages) > sum(have.pkgs)){
  install.packages(required.packages[!have.pkgs])
}

for(pkg in required.packages){
  library(pkg, character.only = TRUE, quietly=TRUE, warn.conflicts=FALSE)
}

# Make sure we're in the right working directory:  "Code"
if("Code" %in% list.files()){
  setwd("Code")
}

if(!("Code" %in% list.files("..")))
{
  stop
}
```


## Problem Definition: Final Project #2     
        
### Project 9.4.2 (page 376): Camparison of 401K VS Social Security for Retirement
        
Retirement and Social Security. Should US citizens build their own retirement through 401Ks or use the current Social Security program?  Build models to be able to compare these systems and provide decisions that can help someone to plan a better retirement.     
        
**Problem Statement:**        
This purpose of this project is to determine the best way to maximize savings for retirement. 401K investment returns and Social Security payout from contributions will be analyzed and compared.      
                
## Solution        
        
### Notes and References:     

Average retirement statistics:  

- Average age of retirement is 62 years
- Average length of retirement is 18 years
- http://www.statisticbrain.com/retirement-statistics/


SSA Calculator Page:  

- Avg men live till 84.3 years
- Avg women live till 86.6 years
- http://www.ssa.gov/retire2/otherthings.htm

Life Expectency Tables

- 1991 birth year (22 year old in 2013) is 75.8 years
- http://www.census.gov/compendia/statab/cats/births_deaths_marriages_divorces/life_expectancy.html

Median Income Tables

- Median income of 22 year old in 2013 is $11,288
- https://www.census.gov/hhes/www/income/data/historical/people/
  
Compound Interest Calculator

- http://www.investor.gov/tools/calculators/compound-interest-calculator

SSA Contribution:

- Tax rate for contribution given by supplement13, Table 2.A3
- Maximum tax amoung given by supplement13, Table 2.A4

SSA Benefit:

- Benefit calculation involves three steps:

  1. Worker's previous earnings are recalulated in current terms with inflation adjustment.
  2. Earnings for the highest 35 years are averaged and divided by 12 to obtain AIME (Averaged Indexed Monthly Earnings).
	3. SSA benefit formula applied to AIME to obtain PIA (Primary Insurance Amount).

- PIA formula with FRA at 2015:

	1. 90 percent of the first $826 of his/her average indexed monthly earnings, plus       
	2. 32 percent of his/her average indexed monthly earnings over $826 and through $4,980, plus      
	3. 15 percent of his/her average indexed monthly earnings over $4,980   
	- http://www.ssa.gov/oact/cola/piaformula.html

- Benefit payouts are adjusted for inflation according to CPI.
- Spousal benefit can also be claimed.
- Taxation of SSA benefits can occur when withdrawing a large amount from IRA due to RMD at 70.5 ("tax torpedo").

- Effect of Early or Delayed Retirement on Retirement Benefits (yearly),  

	Year of Birth	|FRA	| Credit	|62	|63	|64	|65	|66	|67	|70
  --------------|-----|---------|---|---|---|---|---|---|--
	1960 and later| 67 	|8        |	70|75 |	80| 86$\frac{2}{3}$ |93$\frac{1}{3}$ |100 |124

	- http://www.ssa.gov/OACT/ProgData/ar_drc.html

- Average monthly benefit in 2012,  

	Total Number	| ERA (Early retirement)	| FRA (Full Retirement Age)
  ----|-------|---
	$1,147.78	|$1,147.78	|	$1,577.00

	- supplement13, Table 6.B3

- Extrapolate average monthly benefit from 2012,

	Age	|62	|63	|64	|65	|66	|67	|70
  ----|---|---|---|---|---|---|---
	Benefit	|1103.90|	1182.75|1261.60|1366.62|1471.81|1577.00|	1955.48

401K Contribution:

- Tax-deferring contributions until retirement.
- Amount contributed to 401k by retirement without accounting for compound growth

  Age	|62	|63	|64	|65	|66	|67	|70
  ----|---|---|---|---|---|---|---
  401K|690500 |708000 |725500 |743000 |760500 |778000 |795500

  + http://blog.personalcapital.com/financial-planning-2/average-401k-balance-age

- Growth of 401k determined by investment choices.
	- Average of 7% can be expected.
	  + http://www.thesimpledollar.com/where-does-7-come-from-when-it-comes-to-long-term-stock-returns/


401K Benefit:

- Withdrawl without penalties start at 59.5 years.
- RMD (required minimum distribution) starts at either 70.5 or on retirement.   
  + http://www.irs.gov/Retirement-Plans/Plan-Participant/Employee/Retirement-Topics-Required-Minimum-Distributions-%28RMDs%29
- During retirement age, the intended purpose of a 401k was to purchase life annuities.
  However, average payouts for life annuities have been decreasing.
- Savings could be siphoned off by commissions and management fees.
          
### Prepare Data

```{r data, echo=FALSE}
# Consider the age bracket 62 yrs to 70 yrs
age <- c(62, 63, 64, 65, 66, 67, 68, 69, 70)

# SSA - Average montly benefit from 2012
#   supplement13, Table 6.B3 
#   and http://www.ssa.gov/OACT/ProgData/ar_drc.html
mth_amt_rt <- c(1103.9, 1182.75,  1261.6,	1366.62, 1471.81,	1577, 1703.16, 1829.32, 1955.48)
ssa_benefit <- data.frame(age, mth_amt_rt)

# 401k - Base mount accumulated from contribution
#   https://blog.personalcapital.com/financial-planning-2/average-401k-balance-age/
tot_low <- c(690500, 708000, 725500, 743000,	760500, 778000, 795500, 813000, 830500)
# 401k - Balance with 7% compound interest
#   http://www.investor.gov/tools/calculators/compound-interest-calculator
tot_high <- c(3613402, 3883840, 4173209, 4482834,  4814132, 5168621, 5547925, 5953779, 6388044)
k_Balance <- data.frame(age, tot_low, tot_high)

### Assuming life expectancy of 85.45 for modeling:

# Calculate SSA Total Benefit with Average Death at 85.45
ssa_benefit$tot_amt <- (85.45-ssa_benefit$age) * 12 * ssa_benefit$mth_amt_rt

# Calculate 401k Monthly Benefit with Base Balance
k_Balance$mth_low <- k_Balance$tot_low / (85.45-k_Balance$age) / 12
# Calculate 401k Monthly Benefit with 7% Increase
k_Balance$mth_high <- k_Balance$tot_high / (85.45-k_Balance$age) / 12

cdf <- merge(ssa_benefit,k_Balance,by="age")
names(cdf) <- c("Age","SSA_Monthly_Rate","Total_SSA_Benefit","Total_401k_Low","Total_$01k_High","Monthly_$01k_low","Monthly_401k_High")

cat("Prepared Data:\n")
cdf
```

### Comparison of SSA and 401k return.

Assumptions for comparison:

- Start work at age 22.
- Average life expectency till 85.45 for those reaching 65 in 2014.
- Use average SSA benefit from 2012
- Start contributing to 401k from age 22 with contributions the same amount as SSA contribution.
- Use 7% compound growth for 401k contributions.


```{r echo=FALSE}
# 401k Ratios of Benefit To Contribution According to Age
ratio_401k <- k_Balance$tot_high / k_Balance$tot_low

plot(k_Balance$age, ratio_401k, main="401k Ratios of Benefit To Contribution According to Age", xlab="Age", ylab="Benefit ratio")
```

### Modeling:       
        
Problem:

- When to start retirement?
- When to start withdrawing from SSA?


Assumptions for modeling problem:

- Start work at age 22.
- Average life expectency till 85.45 for those reaching 65 in 2014.
- Using average SSA benefit from 2012
- Start contributing to 401k from age 22 with an inital contribution of $8,000 and $17,500 every year after.
- Use 7% compound growth for 401k contributions.


Models: Optimization of Discrete Models with Simplex Method      

```{r echo=FALSE}
# Combined benefit when withdrawing 401k at 62 and delaying SSA
tot_amt <- ssa_benefit$tot_amt + k_Balance[k_Balance$age==62,]$tot_high
mth_amt <- tot_amt / (85.45 * 12)
combin <- data.frame(age, mth_amt, tot_amt)

# Combined benefits from SSA and 401k
plot(combin$age, combin$mth_amt,  main="Combined benefit when withdrawing 401k at 62 and delaying SSA", xlab="Age", ylab="Monthly Amount")

```

```{r echo=FALSE}
# Combined benefit when contributing to 401k until start of SSA
tot_amt <- ssa_benefit$tot_amt + k_Balance$tot_high
mth_amt <- tot_amt / (85.45 * 12)
combin_401contrib <- data.frame(age, mth_amt, tot_amt)

# Combined benefits from SSA and 401k
plot(combin_401contrib$age, combin$mth_amt,  main="Combined benefit when contributing to 401k until start of SSA", xlab="Age", ylab="Monthly Amount")
```
          
## Conclusion 
        
- The best way to maximize savings for retirement is to maximize contributions into a 401K account as early as possible on the planning horizon and to contribute as long as possible.
- Make use of the 401K account earlier, at the start or before retirement.
- Delay the start of SSA benefit payouts as long as possible (up to age 70) to maximize the payout amount throughout retirement.

The contribution and returns on Social Security, 401K, and any other retirement account needs to be considered in combination to better plan for retirement.